---
title: "Report_StormData.Rmd"
author: "Sabin Khadka"
date: "August 21, 2015"
output: html_document
---

# Exploratory data analysis from U.S National Oceanic and Atmospheric Administration's storm database
## Synopsis
In current report, we report the impact of weather on health-related impact on general population and property damages. The data was exported from U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. The data contained characterisctics of major storms and weather events in the U.S, their occurence and estimates of fatalities, injuries and property damage. There were 935 unique events reported. All these events were grouped and recoded to 13 common event types. Here we report sum of health related damage (both injuries and fatalities) and property damage (both crop and property) by events.

## Data analysis
Load all the necessary R libraries.
```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(reshape2)
```
### Data reading
Here we parse and get only subset of columns from data that are related to health, property and events. 
```{r, echo=TRUE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(cache=TRUE)
data <- read.csv(bzfile("repdata_data_StormData.csv.bz2"), header=T, sep=",")
col_interest <- which(colnames(data) %in% c("BGN_DATE", "STATE", "EVTYPE", "INJURIES", "FATALITIES","PROPDMG","PROPDMGEXP","CROPDMG","CROPDMGEXP") )
data <- data[,col_interest]
```
The data in the extracted subset are ```colnames(data)```

To exclude data from the year with less data, we checked number of data present per year.
```{r, echo=TRUE, message=FALSE, warning=FALSE}
data$year = as.numeric(format(as.Date(data$BGN_DATE, format = "%m/%d/%Y"), "%Y"))
nyear <- data$year
n_obs.year <- as.data.frame(table(nyear))
```
Number of data each year ```n_obs.year```

With the help from table we chose year 1989 as our cut-off and only the data from year > 1988 were considered for the analysis.
```{r, echo=TRUE, message=FALSE, warning=FALSE}
data1 <- subset(data, data$year > 1988)
```
We recode the events type to 13 common event types.
```{r, echo=TRUE, message=FALSE, warning=FALSE}
eventsTYPE <- unique(data$EVTYPE)
evTABLE <- matrix(unlist(eventsTYPE), ncol = 5, byrow = TRUE)
totatEventType <- length(eventsTYPE)
ptrn1 <- grepl("burst|depression|drizzle|fog|precipitation|percip|rain|wet|wall cloud", 
               data1$EVTYPE, ignore.case = TRUE)
ptrn2 <- grepl("hail", data1$EVTYPE, ignore.case = TRUE)
ptrn3 <- grepl("hurricane|typhoon|storm|wind|wnd", data1$EVTYPE, ignore.case = TRUE)
ptrn4 <- grepl("erosion|slide|slump", data1$EVTYPE, ignore.case = TRUE)
ptrn5 <- grepl("dry|drought|heat|hot|record temperature|record high|thermia|temperature record|warmth|warm", 
               data1$EVTYPE, ignore.case = TRUE)
ptrn6 <- grepl("avalanche|blizzard|chill|cold|cool|ice|frost|freeze|freezing|glaze|icy|snow|sleet|winter|wintry|wintery", 
               data1$EVTYPE, ignore.case = TRUE)
ptrn7 <- grepl("dam break|flood|fld|surf|swells", data1$EVTYPE, ignore.case = TRUE)
ptrn8 <- grepl("current|drowning|high water|marine|seas|tide|tsunami|wave", data1$EVTYPE, ignore.case = TRUE)
ptrn9 <- grepl("dust|saharan", data1$EVTYPE, ignore.case = TRUE)
ptrn10 <- grepl("lightning|tstm|thunderstorm", data1$EVTYPE, ignore.case = TRUE)
ptrn11 <- grepl("funnel|spout|tornado|whirlwind", data1$EVTYPE, ignore.case = TRUE)
ptrn12 <- grepl("fire|smoke", data1$EVTYPE, ignore.case = TRUE)
ptrn13 <- grepl("volcanic", data1$EVTYPE, ignore.case = TRUE)
# Assign new event code
data1$NewEVCODE <- NA       
data1$NewEVCODE[ptrn1] <- "Rain-related"
data1$NewEVCODE[ptrn2] <- "Hail"
data1$NewEVCODE[ptrn3] <- "Wind-related"
data1$NewEVCODE[ptrn4] <- "Soil-related"
data1$NewEVCODE[ptrn5] <- "Heat-related"
data1$NewEVCODE[ptrn6] <- "Snow/Ice"
data1$NewEVCODE[ptrn7] <- "Flood/Water-related"
data1$NewEVCODE[ptrn8] <- "Sea-related"
data1$NewEVCODE[ptrn9] <- "Dust/Saharan winds"  
data1$NewEVCODE[ptrn10] <- "Thunderstorm/Lightning"
data1$NewEVCODE[ptrn11] <- "Tornado"
data1$NewEVCODE[ptrn12] <- "Fire-related"
data1$NewEVCODE[ptrn13] <- "Volcano"
data1$NewEVCODE <- as.factor(data1$NewEVCODE)
```
Since some of the values of the damages were expressed in alphabets, we change them to numeric form. Also, we derive total damage amount by adding property damage and crop damage.
```{r, echo=TRUE, message=FALSE, warning=FALSE}
convertStr2power <- function(x) {
    if (is.numeric(x)) {
        x <- x
    }
    else if (grepl("h|H", x, ignore.case=T)) {
        x <- 2
    }
    else if (grepl("k|K", x, ignore.case=T)) {
        x <- 3
    }
    else if (grepl("m|M", x, ignore.case=T)) {
        x <- 6
    }
    else if (grepl("b|B", x, ignore.case=T)) {
        x <- 9
    }
    else {
        x <- 0
    }
}

Convert_amt <- function(amt, power){
    power1 <- convertStr2power(power)
    if(is.numeric(amt)){
        tot_amt <- amt * 10^power1
    }
    else if(!is.numeric(amt)){
        tot_amt <- 0
    }
    tot_amt
}

addTWO <- function(x,y){
    tot <- x+y
    tot
}

data1$PROP_DMG_amt <- 0
data1$CROP_DMG_amt <- 0
data1$Total_DMG_amt <- 0
data1$PROP_DMG_amt <- mapply(Convert_amt, data1$PROPDMG, data1$PROPDMGEXP)
data1$CROP_DMG_amt <- mapply(Convert_amt, data1$CROPDMG, data1$CROPDMGEXP)
data1$Total_DMG_amt <- mapply(addTWO, data1$PROP_DMG_amt, data1$CROP_DMG_amt)
```
We calculated the total number of fatalities, injuries and total damage (in $ amount).
```{r, echo=TRUE, warning=FALSE}
Fatal_summary <- aggregate(data1$FATALITIES, by=list(data1$NewEVCODE), FUN=sum)
Injury_summary <- aggregate(data1$INJURIES, by=list(data1$NewEVCODE), FUN=sum)
Amount_summary <- aggregate(data1$Total_DMG_amt, by=list(data1$NewEVCODE), FUN=sum)
```
Here are the codes for the figures generated from the analysis.
```{r, echo=TRUE, warning=FALSE}
merge_result <- Fatal_summary
colnames(merge_result) <- c("Event_Types","Total_Fatalities")
merge_result[, "Total_Injury"] <- Injury_summary$x
merge_result[, "Total_Damage_amt"] <- Amount_summary$x
data2 <- merge_result[,1:3]
mdata2 <- melt(data2)
```
```{r, echo=TRUE, message=FALSE, warning=FALSE}
png("./figure1.png")
p1 <- ggplot(data=mdata2, aes(x=Event_Types, y=value, fill=variable)) + geom_bar(stat="identity") +
    theme(axis.text.x = element_text(size=10, face = "bold",angle = 45, hjust =1))
p1 <- p1 +ggtitle("Total Health related ") + xlab("Event Types") + ylab("Health related Injuries/Fatalities")
print(p1)
dev.off()
```
Figure 1 below shows impact of various event types on population health
```{r, echo=FALSE, fig.height=12, fig.width=8}
plot(p1)
```
```{r, echo=TRUE, message=FALSE, warning=FALSE}
png("./figure2.png")
p2 <- ggplot(data=merge_result, aes(x=Event_Types, y=Total_Damage_amt)) + geom_bar(fill="#DD8888",color="red",stat="identity") +
    theme(axis.text.x = element_text(size=10, face = "bold",angle = 45, hjust =1))
p2 <- p2 +ggtitle("Total loss in $") + xlab("Event Types") + ylab("Loss in $")
print(p2)
dev.off()
```
Figure 2 below shows impact of various event types on economy
```{r, echo=FALSE, fig.height=12, fig.width=8}
plot(p2)
```
### Results
Here we see that the Tornado related events to have caused most health related injuries and fatalities and Flood/water-related events to have caused crop & property damage. Additionally, Wind-related events also caused significant crop & property damage.